The Quartzite Conundrum: Non-destructive analysis of Bronze Age quartzite points from Västerbotten, Sweden

### 2023-01-26 analysis/paper/paper.Rmd
Updated manuscript

---
title: "The Quartzite Conundrum: Non-destructive analysis of Bronze Age quartzite points from Västerbotten, Sweden"
author:
  - Mattias Sjölander:
      email: mattias.sjolander@umu.se
      institute: [UmU]
      correspondence: true
institute:
  - UmU: Umeå University
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
    bookdown::word_document2:
      fig_caption: yes
      reference_docx: "../templates/template.docx" 
      pandoc_args:
      - --lua-filter=../templates/scholarly-metadata.lua
      - --lua-filter=../templates/author-info-blocks.lua
      - --lua-filter=../templates/pagebreak.lua
mainfont: Times New Roman
bibliography: references.bib
csl: "../templates/archaeological-and-anthropological-sciences.csl"
abstract: |
  Northern Fennoscandia represents a geologically complex region affected by both glacial and postglacial processes. Provenance of locally available geological material thus becomes complex and fraught with uncertainty. This leads to a severe lack of methodological development in the characterization of these same materials in an archaeological context. Quartzite has been a central material type utilized among hunter-gatherers in Northern Sweden around the 4 000 – 2 000 BP period and is thus important to the understanding of raw material procurement and material flow within the region. 125 points and preforms made from quartz/quartzite material were sampled from 46 archaeological sites located along the upper Ångerman river valley in Västerbotten, Sweden. The material has been analysed non-destructively using three separate spectroscopic instrumentations (Near-infrared, Raman, X-Ray Fluorescence). The data was evaluated with exploratory analysis using Principal Component Analysis. The results demonstrate that detectable differences exist within the dataset for each instrumentation. A number of features identified, such as the presence of graphite, can likely be related to diagenetic/paragenetic differences in the material. Not only should these results aid in future surveys of geological sources in the region, but they will also aid in the construction of predictive models for classifying quartzite.  
keywords: |
  Lithics; Provenance; Quartzite; Spectroscopy; Sweden
highlights: |
  These are the highlights. 
---

@@

```{r setup, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  warning = FALSE,
  error = TRUE,
  message = FALSE,
  echo = FALSE,
  comment = "#>",
  fig.path = "../figures/",
  dpi = 300,
  encoding = "UTF-8"
)
```

@@

# Introduction

@@

# Aims

@@

# Material

@@

# Methods

@@

## Instrumentaion and pre-processing

@@

### Near Infrared
NIRS measures the absorbance of light within the near infrared region (780 nm – 2 500 nm) of the electromagnetic spectrum. Based on the theory of molecular vibration and excitation [@hunt77], by exposing a geological material to electromagnetic radiation it is possible to generate overtone vibrations which can be related to certain molecular groups. The most common group capable of generating these overtones is the OH group, which in quartz-based materials typically relates to molecular water and metal-OH combinations [@hunt77 p.508].

@@

### Raman
Raman spectroscopy shares some similarities with NIRS in that they both can be used to infer molecular structure based on their vibrational properties within the near-infrared spectrum [@nafie01]. However, where NIRS measures light absorbance and generate overtones of molecular vibration, Raman spectroscopy is based on the *Raman scattering* effect. By exposing a molecule to radiation with a monochromatic laser it is possible to induce a dipole moment in the molecule which generates the Raman effect. The subsequent light that is scattered contains both Rayleigh and Raman scatter, the former at a frequency of incident radiation and the latter at a shifted frequency [@dubessy_etal12]. The spectrometer will measure the Raman scatter and generate a spectrum where the spectral line has shifted compared to the incident Rayleigh frequency. As the structure of a molecule will inform its vibrational mode, the point at which these shifts occur within the infrared region varies depending on the material [@smith_carabatos-nedelec01].

@@

### X-Ray Fluourescence
XRF spectroscopy, and its capabilities for quantifying chemical elements in different materials, has made it a well-established method in petrographic characterization in archaeology [@hall60; @shackley11a]. Moseley's law [@moseley1913] recognizes that the energy of the fluorescence produced when a material is irradiated with high energy particles is characteristic to certain atoms. This is what enables the spectral features in XRF spectrometry to be related to certain elements. Quantification based on these features is complex, however, and a number of methods have been developed as a result [@sitko_zawisa12 pp. 143-159]. In a non-destructive context, where the sample is not homogeneous, it is necessary to correct for matrix effects that are caused by the presence of other elements in a sample. One way to do so is calibration using pure-element standards, with which a calibration constant can be calculated.

@@

## PCA

@@

# Results

@@

## NIR - near infrared range
```{r nir-sample-count, results='hide'}
#Select all dark samples
points.d.tot <-
  metadata.csv %>%
  filter(site_id == "Vilhelmina 1069" | site_id == "Vilhelmina 109" | site_id == "Vilhelmina 112" | site_id == "Vilhelmina 1124" | site_id == "Vilhelmina 1127" | 
           site_id == "Vilhelmina 114" | site_id == "Vilhelmina 115" | site_id == "Vilhelmina 117" | site_id == "Vilhelmina 118" | site_id == "Vilhelmina 1254" | 
           site_id == "Vilhelmina 216" | site_id == "Vilhelmina 235" | site_id == "Vilhelmina 240" | site_id == "Vilhelmina 245" | site_id == "Vilhelmina 252" | 
           site_id == "Vilhelmina 263" | site_id == "Vilhelmina 335" | site_id == "Vilhelmina 356" | site_id == "Vilhelmina 399" | site_id == "Vilhelmina 411" | 
           site_id == "Vilhelmina 419" | site_id == "Vilhelmina 439" | site_id == "Vilhelmina 444" | site_id == "Vilhelmina 450" | site_id == "Vilhelmina 458" | 
           site_id == "Vilhelmina 539" | site_id == "Vilhelmina 542" | site_id == "Vilhelmina 611" | site_id == "Vilhelmina 619" | site_id == "Vilhelmina 636" | 
           site_id == "Vilhelmina 637" | site_id == "Vilhelmina 643" | site_id == "Vilhelmina 769" | site_id == "Vilhelmina 949" | site_id == "Vilhelmina 95" | 
           site_id == "Åsele 101" | site_id == "Åsele 107" | site_id == "Åsele 115" | site_id == "Åsele 117" | site_id == "Åsele 119" | site_id == "Åsele 129" | 
           site_id == "Åsele 182" | site_id == "Åsele 188" | site_id == "Åsele 393" | site_id == "Åsele 56" | site_id == "Åsele 91" | site_id == "Åsele 92" | 
           site_id == "Åsele 99", 
         type == "Point" | type == "Point fragment" | type == "Preform", 
         material == "Brecciated quartz" | material == "Quartz" | material == "Quartzite") %>% 
  filter(hue == "Dark")

@@

The loading lines for the PCA can be seen in Appendix \@ref(fig:nir-pca-loadings). The first principal component (PC) shows a strong trend (`r nir.pc1var` %) that seems to largely correlate with the absorbance of the material (fig. \@ref(fig:nir-pca-fig)). This can be seen in the trend of darker material clustering towards the positive side of PC 1, with lighter towards the negative. A slight difference in the clustering of material with a light hue can also be seen along PC 2 (`r nir.pc2var` %). Objects with a  white or light grey hue gravitate towards the bottom left of the score plot, while material that showcases some translucence cluster towards the top left. There are objects that break this trend, however, such as sample 255 in the bottom right of fig. \@ref(fig:nir-pca-fig):A.

@@

## Raman
There are two distinguishing features present in the Raman data (fig. \@ref(fig:raman-spectra-summary) & fig. \@ref(fig:raman-peak-ident)), one which can be related to quartz and another that may indicate the presence of carbon (C) structures. The three initial bands at ~ 128 cm^-1^, ~ 200 cm^-1^ and ~ 460 cm^-1^, are some of the more intense quartz bands [@zhong_etal21], with the strongest peak occurring at 460 cm^-1^ [@gillet_etal90].

@@

The loading lines for the PCA can be seen in Appendix \@ref(fig:raman-pca-loadings). The PCA model (fig. \@ref(fig:raman-pca)) shows a clear trend along PC 1 (`r raman.pc1var` %) related to the quartz and D/G peaks. Samples where the D/G peaks dominate in terms of intensity cluster towards the positive side of PC 1, with those that completely lack or have weaker D/G bands on the opposite side. Similarly, along PC 2 (`r raman.pc2var` %) the samples that have higher intensity D/G peaks gravitate towards the top of the score plot. Worth noting are two dark samples that completely lack D/G peaks, placing them among the light material in the left side of the plot. One of these is a brecciated quartz sample with a dark yellow tint, and thus more a sign of the difficulties with assigning a generic hue to the material. The second sample, however, stands out from the rest of the material. It is a point preform made from a grey material with a rough texture, and possible layering, with larger quartz crystals. It does not fit with the "expected" image of a quartzite, and may be a case of mischaraterization. The addition of PC 3 (`r raman.pc3var` %) provides no new information, but rather separates out those samples with a higher level of noise.

@@

## XRF

@@

As can be seen in the score plot of PC 1-2 (fig. \@ref(fig:xrf-pca-a)), the three K-means clusters remain separated with next to no overlap. Along PC 1 cluster 3 is marked by a higher ~26~Fe and ~40~Zr content on average, as well as a ~14~Si content lower than 50%. It mainly features samples with a dark hue, with the few samples from the light group being grey in color and arguably somewhere in between what might be considered a "dark" or "light" hue. Cluster 2 is separated out along PC 2, and includes only samples with a ~14~Si content above 90%, while ~26~Fe and ~40~Zr is low or below the LOD. Cluster 1 includes the main bulk of the light group, as well as all of the *brecciated quartz* samples. As in the case of cluster 3, apart from 1 outlier, the remaining samples in cluster 1 have a ~14~Si content below 50%. Except for ~22~Ti, ~26~Fe and ~40~Zr, the cluster features comparable values to cluster 3. Where it stands out is in the large amount of samples with ~38~Sr readings. ~38~Sr is a lithophile element that can substitute for ~20~Ca in sulphates, feldspars, mica and clays, and has been the focus of study of carbonate diagenesis/paragenesis [@ahijado_etal05;@banner95]. 
-->
The PCA was run twice, once with the > 90% ~14~Si samples included and a second time without them. The loadings can be seen in Appendix \@ref(fig:xrf-pca-loadings-scree). The initial PCA (model A), with the full dataset included, resulted in 3 PCs with an eigenvalue above 1, and a cumulative explained variance of `r xrf_var`%. K-means cluster analysis was performed on the PCA data in order identify potential groups with similar elemental composition. Using the elbow graph method 3 clusters were deemed suitable for the analysis.

@@

# Discussion

@@

# Conclusion

@@

# Acknowledgements

I would like to thank Professor Paul Geladi, as well as my supervisor Dr. Johan Linderholm, for valuable insights and comments on the methodological aspects of this paper.

@@

# References

@@

# Appendix A

@@

```{r nir-pca-loadings, fig.width=20, fig.height=20, fig.cap= "PCA loadings for the first 3 principal components for the NIR data. ", cache=TRUE, cache.extra=tools::md5sum(here::here("analysis", "paper", "004-nir-pca-loadings.R")), echo = FALSE}
quietly(source(here::here("analysis", "paper", "004-nir-pca-loadings.R"), encoding = "UTF-8"))
knitr::include_graphics(here::here("analysis", "figures", "004-nir-pca-loadings.png"), error = FALSE)
```

@@

```{r raman-pca-loadings, fig.cap= "PCA loadings for the first 3 principal components for the Raman data", cache=TRUE, cache.extra=tools::md5sum(here::here("analysis", "paper", "007-raman-pca-loadings.R")), echo = FALSE}
quietly(source(here::here("analysis", "paper", "007-raman-pca-loadings.R"), encoding = "UTF-8"))
knitr::include_graphics(here::here("analysis", "figures", "007-raman-pca-loadings.png"), error = FALSE)
```

@@

```{r xrf-pca-loadings-scree, fig.cap= "PCA Loadings and eigenvalues for the XRF data. a) eigenvalues and loadings for the full dataset, b) eigenvalues and loadings with samples measuring > 90% ~14~Si excluded.", echo = FALSE}
knitr::include_graphics(here::here("analysis", "figures", "009-xrf-pca-load-scree.png"), error = FALSE)
```

@@

### Colophon

>>>>

---
title: "The Quartzite Conundrum: Non-destructive analysis of Bronze Age quartzite points from Västerbotten, Sweden"
author:
  - Mattias Sjölander:
      email: mattias.sjolander@umu.se
      institute: [UmU]
      correspondence: true
  - Johan Linderholm:
      email: johan.linderholm@umu.se
      institute: [UmU]
      correspondence: false
  - Paul Geladi:
      email: paul.geladi@slu.se
      institute: [SLU]
      correspondence: false
institute:
  - UmU: Department of Historical, Philosophical and Religious studies, Umeå University, Sweden
  - SLU: Department of Forest Biomaterials and Technology, Swedish University of Agricultural Sciences, Sweden 
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
    bookdown::word_document2:
      number_sections: false
      fig_caption: yes
      reference_docx: "../templates/template.docx" # Insert path for the DOCX file
      pandoc_args:
      - --lua-filter=../templates/scholarly-metadata.lua
      - --lua-filter=../templates/author-info-blocks.lua
      - --lua-filter=../templates/pagebreak.lua
header-includes:
    \newcommand{\beginsupplement}{\setcounter{table}{0}  \renewcommand{\thetable}{S\arabic{table}} \setcounter{figure}{0} \renewcommand{\thefigure}{S\arabic{figure}}}
bibliography: references.bib
csl: "../templates/archaeological-and-anthropological-sciences.csl"
abstract: |
  Northern Fennoscandia represents a geologically complex region affected by both glacial and postglacial processes. Provenance of locally available geological material thus becomes complex and fraught with uncertainty. This leads to a severe lack of methodological development in the characterization of these same materials in an archaeological context. Quartzite has been a central material type utilized among hunter-gatherers in Northern Sweden around the 4 000 – 2 000 BP period and is thus important to the understanding of raw material procurement and material flow within the region. 125 points and preforms made from quartz/quartzite material were sampled from 46 archaeological sites located along the upper Ångerman river valley in Västerbotten, Sweden. The material has been analysed non-destructively using three separate spectroscopic instrumentations (Near-infrared, Raman, X-Ray Fluorescence). The data was evaluated with exploratory analysis using Principal Component Analysis. The results demonstrate that detectable differences exist within the dataset for each instrumentation. A number of features identified, such as the presence of graphite, can likely be related to diagenetic/paragenetic differences in the material. Not only should these results aid in future surveys of geological sources in the region, but they will also aid in the construction of predictive models for classifying quartzite.  
keywords: |
  Lithics; Provenance; Quartzite; Spectroscopy; Sweden
highlights: |
  These are the highlights. 
---

@@

```{r setup, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  warning = FALSE,
  message = FALSE,
  echo = FALSE,
  comment = "#>",
  fig.path = "../figures/",
  dpi = 300,
  encoding = "UTF-8"
)
```

@@

# Introduction {-}

@@

# Aims {-}

@@

# Material {-}

@@

# Methods {-}

@@

## Instrumentaion and pre-processing {-}

@@

### Near Infrared {-}
NIRS measures the absorbance of light within the near infrared region (780 nm – 2 500 nm) of the electromagnetic spectrum. Based on the theory of molecular vibration and excitation [@hunt77], by exposing a geological material to electromagnetic radiation it is possible to generate overtone vibrations which can be related to certain molecular groups. The most common group capable of generating these overtones is the OH group, which in quartz-based materials typically relates to molecular water and metal-OH combinations [@hunt77 p.508].

@@

### Raman {-}
Raman spectroscopy shares some similarities with NIRS in that they both can be used to infer molecular structure based on their vibrational properties within the near-infrared spectrum [@nafie01]. However, where NIRS measures light absorbance and generate overtones of molecular vibration, Raman spectroscopy is based on the *Raman scattering* effect. By exposing a molecule to radiation with a monochromatic laser it is possible to induce a dipole moment in the molecule which generates the Raman effect. The subsequent light that is scattered contains both Rayleigh and Raman scatter, the former at a frequency of incident radiation and the latter at a shifted frequency [@dubessy_etal12]. The spectrometer will measure the Raman scatter and generate a spectrum where the spectral line has shifted compared to the incident Rayleigh frequency. As the structure of a molecule will inform its vibrational mode, the point at which these shifts occur within the infrared region varies depending on the material [@smith_carabatos-nedelec01].

@@

### X-Ray Fluourescence {-}
XRF spectroscopy, and its capabilities for quantifying chemical elements in different materials, has made it a well-established method in petrographic characterization in archaeology [@hall60; @shackley11a]. Moseley's law [@moseley1913] recognizes that the energy of the fluorescence produced when a material is irradiated with high energy particles is characteristic to certain atoms. This is what enables the spectral features in XRF spectrometry to be related to certain elements. Quantification based on these features is complex, however, and a number of methods have been developed as a result [@sitko_zawisa12 pp. 143-159]. In a non-destructive context, where the sample is not homogeneous, it is necessary to correct for matrix effects that are caused by the presence of other elements in a sample. One way to do so is calibration using pure-element standards, with which a calibration constant can be calculated.

@@

## PCA {-}

@@

# Results {-}

@@

## NIR - near infrared range {-}
```{r nir-sample-count, results='hide'}
#Select all dark samples
points.d.tot <-
  metadata.csv %>%
  filter(site_id == "Vilhelmina 1069" | site_id == "Vilhelmina 109" | site_id == "Vilhelmina 112" | site_id == "Vilhelmina 1124" | site_id == "Vilhelmina 1127" | 
           site_id == "Vilhelmina 114" | site_id == "Vilhelmina 115" | site_id == "Vilhelmina 117" | site_id == "Vilhelmina 118" | site_id == "Vilhelmina 1254" | 
           site_id == "Vilhelmina 216" | site_id == "Vilhelmina 235" | site_id == "Vilhelmina 240" | site_id == "Vilhelmina 245" | site_id == "Vilhelmina 252" | 
           site_id == "Vilhelmina 263" | site_id == "Vilhelmina 335" | site_id == "Vilhelmina 356" | site_id == "Vilhelmina 399" | site_id == "Vilhelmina 411" | 
           site_id == "Vilhelmina 419" | site_id == "Vilhelmina 439" | site_id == "Vilhelmina 444" | site_id == "Vilhelmina 450" | site_id == "Vilhelmina 458" | 
           site_id == "Vilhelmina 539" | site_id == "Vilhelmina 542" | site_id == "Vilhelmina 611" | site_id == "Vilhelmina 619" | site_id == "Vilhelmina 636" | 
           site_id == "Vilhelmina 637" | site_id == "Vilhelmina 643" | site_id == "Vilhelmina 769" | site_id == "Vilhelmina 949" | site_id == "Vilhelmina 95" | 
           site_id == "Åsele 101" | site_id == "Åsele 107" | site_id == "Åsele 115" | site_id == "Åsele 117" | site_id == "Åsele 119" | site_id == "Åsele 129" | 
           site_id == "Åsele 182" | site_id == "Åsele 188" | site_id == "Åsele 393" | site_id == "Åsele 56" | site_id == "Åsele 91" | site_id == "Åsele 92" | 
           site_id == "Åsele 99", 
         type == "Point" | type == "Point fragment" | type == "Preform", 
         material == "Brecciated quartz" | material == "Quartz" | material == "Quartzite") %>% 
  filter(hue == "Dark")

@@

The loading lines for the PCA can be seen in [Appendix A](#appendix-a). The first principal component (PC) shows a strong trend (`r nir.pc1var` %) that seems to largely correlate with the absorbance of the material (fig. \@ref(fig:nir-pca-fig)). This can be seen in the trend of darker material clustering towards the positive side of PC 1, with lighter towards the negative. A slight difference in the clustering of material with a light hue can also be seen along PC 2 (`r nir.pc2var` %). Objects with a  white or light grey hue gravitate towards the bottom left of the score plot, while material that showcases some translucence cluster towards the top left. There are objects that break this trend, however, such as sample 255 in the bottom right of fig. \@ref(fig:nir-pca-fig):A.

@@

## Raman {-}
There are two distinguishing features present in the Raman data (fig. \@ref(fig:raman-spectra-summary) & fig. \@ref(fig:raman-peak-ident)), one which can be related to quartz and another that may indicate the presence of carbon (C) structures. The three initial bands at ~ 128 cm^-1^, ~ 200 cm^-1^ and ~ 460 cm^-1^, are some of the more intense quartz bands [@zhong_etal21], with the strongest peak occurring at 460 cm^-1^ [@gillet_etal90].

@@

The loading lines for the PCA can be seen in [Appendix A](#appendix-a). The PCA model (fig. \@ref(fig:raman-pca)) shows a clear trend along PC 1 (`r raman.pc1var` %) related to the quartz and D/G peaks. Samples where the D/G peaks dominate in terms of intensity cluster towards the positive side of PC 1, with those that completely lack or have weaker D/G bands on the opposite side. Similarly, along PC 2 (`r raman.pc2var` %) the samples that have higher intensity D/G peaks gravitate towards the top of the score plot. Worth noting are two dark samples that completely lack D/G peaks, placing them among the light material in the left side of the plot. One of these is a brecciated quartz sample with a dark yellow tint, and thus more a sign of the difficulties with assigning a generic hue to the material. The second sample, however, stands out from the rest of the material. It is a point preform made from a grey material with a rough texture, and possible layering, with larger quartz crystals. It does not fit with the "expected" image of a quartzite, and may be a case of mischaraterization. The addition of PC 3 (`r raman.pc3var` %) provides no new information, but rather separates out those samples with a higher level of noise.

@@

## XRF {-}

@@

As can be seen in the score plot of PC 1-2 (fig. \@ref(fig:xrf-pca-a)), the three K-means clusters remain separated with next to no overlap. Along PC 1 cluster 3 is marked by a higher ~26~Fe and ~40~Zr content on average, as well as a ~14~Si content lower than 50%. It mainly features samples with a dark hue, with the few samples from the light group being grey in color and arguably somewhere in between what might be considered a "dark" or "light" hue. Cluster 2 is separated out along PC 2, and includes only samples with a ~14~Si content above 90%, while ~26~Fe and ~40~Zr is low or below the LOD. Cluster 1 includes the main bulk of the light group, as well as all of the *brecciated quartz* samples. As in the case of cluster 3, apart from 1 outlier, the remaining samples in cluster 1 have a ~14~Si content below 50%. Except for ~22~Ti, ~26~Fe and ~40~Zr, the cluster features comparable values to cluster 3. Where it stands out is in the large amount of samples with ~38~Sr readings. ~38~Sr is a lithophile element that can substitute for ~20~Ca in sulphates, feldspars, mica and clays, and has been the focus of study of carbonate diagenesis/paragenesis [@ahijado_etal05;@banner95]. 
-->
The PCA was run twice, once with the > 90% ~14~Si samples included and a second time without them. The loadings can be seen in [Appendix A](#appendix-a). The initial PCA (model A), with the full dataset included, resulted in 3 PCs with an eigenvalue above 1, and a cumulative explained variance of `r xrf_var`%. K-means cluster analysis was performed on the PCA data in order identify potential groups with similar elemental composition. Using the elbow graph method 3 clusters were deemed suitable for the analysis.

@@

# Discussion {-}

@@

# Conclusion {-}

@@

# CRediT authorship contribution statement {-}
Mattias Sjölander: Conceptualization, Data curation, Methodology, Software, Validation, Formal analysis, Resources, Writing - original draft, Visualization, Project administration.

Johan Linderholm: Conceptualization, Methodology

Paul Geladi: Methodology

# Statements and Declarations

# Funding source {-}
This work was supported by

# Acknowledgements {-}

We would like to thank Västerbottens museum (www.vbm.se) for giving us access to their collections, and Susanne Sundström for her assistance with putting the artefacts together. We would also like to acknowledge the XXX consortium for providing access to the instrumentation used in this study.

@@

# References {-}

@@

## Appendix A {#appendix-a}

Collection of loadings for each single instrumentation PCA.

\beginsupplement

@@

```{r nir-pca-loadings, fig.width=20, fig.height=20, fig.cap= "PCA loadings for the first 3 principal components for the NIR model.", cache=TRUE, cache.extra=tools::md5sum(here::here("analysis", "paper", "004-nir-pca-loadings.R")), echo = FALSE}
quietly(source(here::here("analysis", "paper", "004-nir-pca-loadings.R"), encoding = "UTF-8"))
knitr::include_graphics(here::here("analysis", "figures", "004-nir-pca-loadings.png"), error = FALSE)
```

@@

```{r raman-pca-loadings, fig.cap= "PCA loadings for the first 3 principal components for the Raman model", cache=TRUE, cache.extra=tools::md5sum(here::here("analysis", "paper", "007-raman-pca-loadings.R")), echo = FALSE}
quietly(source(here::here("analysis", "paper", "007-raman-pca-loadings.R"), encoding = "UTF-8"))
knitr::include_graphics(here::here("analysis", "figures", "007-raman-pca-loadings.png"), error = FALSE)
```

@@

```{r xrf-pca-loadings-scree, fig.cap= "PCA Loadings and eigenvalues for the XRF data. a) eigenvalues and loadings for the full dataset (model A), b) eigenvalues and loadings with samples measuring > 90% ~14~Si excluded (model B).", echo = FALSE}
knitr::include_graphics(here::here("analysis", "figures", "009-xrf-pca-load-scree.png"), error = FALSE)
```

@@

### Colophon {-}
